---
title: "EDA_week5excercise1"
author: "Lucia M Rodriguez Bravo"
date: "2/27/2020"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Messy Data Activity

## Data extraction
First were calling the packages to be used

```{r}
library(tidyverse)
library(dbplyr)
```

Then, we'll call the deliberately broken dataset 

This dataset contains the following variables: 

netID: A unique identifier for the sampling event (more than one can share)

fishID: A unique identifier for the individual fish (can't be duplicated)

species: Species name

length: Total length (inches to nearest 0.1)

weight: Wet weight (grams to nearest 0.1)

year: Year of capture

# STEP 1: Did it load correctly?
```{r}
InchLake2_Broken<-read.csv("data/InchLake2-Broken.csv")
head(InchLake2_Broken)
#Extra columns? Yes. Column X. We get rid of it.
InchLake2_Broken <- InchLake2_Broken %>% select(-X)
head(InchLake2_Broken)
#It worked! Now, is there any extra rows? Also yes. The last row has a weight registered but nothing else. 
#To remove it:
InchLake2_Broken <- InchLake2_Broken %>%filter(!is.na(fishID)) #any variable besides weight should work

```

# STEP 2: Check data types
```{r}
# First Question: What are the data types in this data set?
sapply(InchLake2_Broken, class)
# Does this look right? No. weight and year should not be a "factor"
```
### Weight first
```{r}
levels(InchLake2_Broken$weight)
#looks like the value "1,070" has a comma where it shouldn't. We create a new dataframe
InchLake_fix<-InchLake2_Broken
InchLake_fix$weight<-as.character(InchLake_fix$weight) #make factors into characters
InchLake_fix %>%filter(is.na(as.numeric(weight))==TRUE) #filter out non-numeric characters
#the mistake, as expected, is "1,070" We could fix it as follows:
InchLake_fix <- InchLake_fix %>%mutate(weight = ifelse(weight=="1,070", 1070, weight))
InchLake_fix$weight<-as.numeric(InchLake_fix$weight)
#alternatively: InchLake_fix$weight<-gsub(",","",InchLake_fix$weight)  
#(Still, a 17 inch bass weighing over 1kg seems suspicious, no?)
range(InchLake_fix$weight)
plot(InchLake_fix$weight)
#did it work?
sapply(InchLake_fix, class)
#it is our highest value, but it's not obviously an outlier
```
### Now year
```{r}
levels(InchLake_fix$year) #there's a 2048, a 2o07 and a 2 008. Must fix each
InchLake_fix$year<-as.character(InchLake_fix$year)
InchLake_fix %>%filter(is.na(as.numeric(year))==TRUE) #note: 2048 doesn't appear because it is a number
InchLake_fix <- InchLake_fix %>%mutate(year = ifelse(year=="2 008", 2008, year))%>%mutate(year = ifelse(year==2048, 2008, year))%>%mutate(year = ifelse(year=="2O07", 2007, year))
InchLake_fix$year<-as.numeric(InchLake_fix$year)
#checking
range(InchLake_fix$year)
plot(InchLake_fix$year)

sapply(InchLake_fix, class)
```


# STEP 3: Check for missing or impossible values
```{r}
#checking the ranges of everything except species
range(InchLake_fix$netID) #looks ok
plot(InchLake_fix$netID) #looks ok

range(InchLake_fix$fishID) #not super useful 
plot(InchLake_fix$fishID)

range(InchLake_fix$length) #there's a minus value, that makes no sense
plot(InchLake_fix$length) #there's also one lone very high value (800)
InchLake_fix <- InchLake_fix %>%mutate(length = ifelse(length=="-8.7", 8.7, length))%>%mutate(length = ifelse(length==800, 8.0, length)) #chose to make our 800 inch fish a 8 inch fish to match best other bluegills
#check it worked
range(InchLake_fix$length)
plot(InchLake_fix$length) 

#Still need to check fishID and species
#Fish ID should be unique per fish so
InchLake_fix %>%count(n_distinct(fishID)) #there's 2 less unique values than there should
InchLake_fix %>% group_by(fishID) %>% filter(n()>1) # fish ID 517 is repeated 3 times. I would have to look at the original data to figure out which is which. We corect accordingly:
InchLake_fix <- InchLake_fix%>%
  mutate(fishID = ifelse(fishID==517 & length==2.2, 516, fishID)) %>% mutate(fishID = ifelse(fishID==517 & length==9.0, 222, fishID))
#Did it work?
InchLake_fix %>%count(n_distinct(fishID)) #Yes

```


# STEP 4: Check for typos and broken factors
```{r}
#The last check that needs to happen: species

levels(InchLake_fix$species) #mistakes on several
InchLake_fix$species<-as.character(InchLake_fix$species)
InchLake_fix <- InchLake_fix%>%
  mutate(species = ifelse(species=="Black Crapie", "Black Crappie", species)) %>% mutate(species = ifelse(species=="bluegill", "Bluegill", species))%>% mutate(species = ifelse(species=="Bluegill ", "Bluegill", species))%>% mutate(species = ifelse(species=="Iowa_Darter ", "Iowa Darter", species))%>% mutate(species = ifelse(species=="Largemout Bass", "Largemouth Bass", species))
InchLake_fix$species<-as.factor(InchLake_fix$species)
#Did it work?
levels(InchLake_fix$species) 
```
##Final check
```{r}
InchLake2_Clean<- read.csv("data/InchLake2-Clean.csv")

all.equal(InchLake2_Clean, InchLake_fix)
```
# B)

